---
title: "What is a model?"
author: "Kate Becker"
date: "2024-08-26"
output: html_document
---

```{r}
library(tidyverse)
library(janitor)
```

A Model is any sort of function that has predictive power 

```{r}
head(mtcars)
```
- Columns of data are called features 

Examining if theres a relationship between the cars fuel efficiency and any of the features 

```{r}
# matrix of scatterplots if producted with pairs()
# lower.panel separate panel functions to be used above diaganol
# clearly nothing in the cyl row look like they help for simple regression modelling
pairs(mtcars[1:7], lower.panel = NULL)
```
- MPG as a function of cycl looks very different from mpg as a function of wt 

```{r}
plot(y = mtcars$mpg, x = mtcars$wt, xlab = "Vehicle Weight", ylab = "Vehicle Fuel Efficiency in Miles per Gallon")
```
From this kind of data format, we can extract a best fit to all the data points and turn this plot into an equation. 

Using a function to model the value were interested in, called a response, against other features in the dataset

```{r}
mt.model <- lm(formula = mpg ~ wt, data = mtcars)
mt.model
```
```{r}
coef(mt.model)[2]
coef(mt.model[1])
```

Now we have predictive power given some kind of input that can give us a value for any number we put in but there may be error in data or observation. 

# Algorithms

An algorithm is a set of steps performed in order 
- Simplest: linear regression, involves putting two points on a plot and then drawing a line between them you get the important parts of the question (slope and intercept) by taking the difference in the coordinates of those points with respect to some origin 
- this gets more complicated with more than two points though so need more complex systems that humans cant do to do this 
- Algorithsm are the underpinnings of simple R code 

# 3 major types of models: 
- Regression models 
- Classification models
- Mixed models (Combo of both)

# Classificaiton models 
- Take input data and arrange it according to a type, class, group, or other discrete output 

# Mixed models
- Start with a regression model and then use output from that to help to classify other types of data and the reverse can be true 
- lm(y~x) "give me the lienar model for the variable y as a function of the feature x"

Function: object that has some processing power, sits inside model
Model: complex object that takes an input parameter and gives an output 
Equation: mathematical representation of a funciton 
Algorithm: set of steps passed into a model for calculating or processing 
- the lm() is a function itself, but also a linear model 
  - calls a series of algorithms to find best values that are then output as a slope and an intercept 
  
  All models are wrong like the mtcars data isnt comparable to data we have now
  
  Coefficient of determination: measure of how close the data fits the model fitted line with values ranging from 0 to 1 
  
- 

```{r}

```

